---
title: "A1"
author: "Rafsaan Sanvir"
date: "2024-06-06"
output: word_document
---

The data file houses.txt contains measurements of house prices (in $1000s)
and living area (in square feet) for 25 houses. Suppose we consider the simple normal
linear regression to describe the relationship between the response Y (price) and the
predictor X (living area). Answer the following data related questions: 


Import dataset and view
```{r}
library(readr)
houses <- read.table("houses.txt", header = TRUE)
head(houses)
```

3a) Plot the data in a scatter plot and comment on your plot.
Solution: 
```{r}
x = houses$living_area
y = houses$price
plot(x, y)
```
The plot above seems to indicate a strong positive linear relationship. 

3b) From now on we will assume that the Normal error regression model we
discussed in class is appropriate for this data. Calculate the least squares regression
line.
Solution: 
```{r}
#manual calculation
b1 = sum((x - mean(x))*(y - mean(y))) / sum((x-mean(x))^2) #by least sq formula
cat("least square est of b1 is", b1, "\n")
b0 = mean(y) - b1 * mean(x)
cat("least square est of b0 is", b0)

fit = lm(y~x)
summary(fit)
```
Note how the least squares estimate matches the b0 and b1 value provided by R 

3c) Conduct a t-test to determine whether or not there is a linear association
between X and Y . Use α = 0.05 State the null and alternative hypotheses.

Solution: H0: B1 = 0 vs HA: B1 != 0 
Note: From the lm summary above, stand error of b1 can be extracted, & n = 25
```{r}
sxx = sum((x - mean(x))^2)
alpha = 0.05
sb1 = summary(fit)$coefficients[2,2]
tstar = (b1 - 0)/sb1
cat("test stat is:", tstar, "\n") 

n = 25
tcritical = qt(1 - alpha/2, n - 2)
cat("critical value is:", tcritical, "\n")

```

Since Test statistic value 14.84 much greater than critical value 2.06, test stat 
falls in rejection region and we reject the null hypothesis. Meaning that a 
significant linear association between x and y exists. 

d) Calculate the power of your test in part (c) above if actually β1 = 0.1.
Assume σ = 44.
Solution: 
```{r}
variance = 44^2
sxx = sum((x - mean(x))^2)
variance_b1 = variance / sxx # by formula 
b1_not = 0.1
delta = abs(b1_not - 0) / sqrt(variance_b1) # by delta formula
power = pt(qt(1 - (alpha / 2), n-2), n-2, ncp = delta, low = F) + 
  pt(-qt(1 - (alpha / 2), n-2 ), n-2, ncp = delta, low = T)
power
```
This means that if b1 = 0.1, the probability would by 0.88 that we would be 
led to conclude with the alternate hypothesis that b1 != 0. 

e) Calculate a 95% confidence interval for β1.
Solution: 
```{r}
confint(fit, level = 0.95)
```
95% CI for b1 is (0.1222, 0.1617).
This means that if we conducted this analysis multiple times with different 
samples from the same population, 95% of the calculated confidence intervals 
would contain the true value of b1. The confidence interval also does not contain
0, meaning a statistically significant relationship exists between living 
area and house price.s 

f) Calculate a 95% cofidence interval for the mean price of houses with a
living area of 2000 square feet.
Solution: 
```{r}
newdata = data.frame(x = 2000)
predict(fit, newdata = newdata, interval = "confidence", level = 0.95)
```
Confidence interval is (346.2627, 360.6736) 

g) Calculate a 95% prediction interval for the price of a new house with a
living area of 2000 square feet.
Solution: 
```{r}
predict(fit, newdata = newdata, interval = "pred", level = 0.95)
```
Prediction interval is (324.0093, 382.927)
We are 95% confident that the average price of houses with a living area of 2000 
square feet is between 346.2627 and 360.6736.

h) Calculate the boundary values of the 95% confidence band for the re-
gression line when the living area is Xh = 2000 square feet.
Solution: Let's use the boundary value formula 
```{r}
xh = 2000
xbar = mean(x)
yhat = fit$fitted.values
SSE  = sum((y - yhat)^2)
MSE = SSE / (n - 2)
s = sqrt(MSE)
xbar = mean(x)
yhath = fit$coefficients[1] + fit$coefficients[2] *xh
syhath = s * sqrt(1/n+((xh-xbar)^2) / sxx)
W = sqrt(2 * qf(1 - alpha, df1 = 2, df2 = n-2))
wh.lower = yhath - W * syhath
wh.upper = yhath + W*syhath
cat("lower value is ", wh.lower, "\n")
cat("upper value is ", wh.upper, "\n")
```

This means we are 95% confident that the true regression line (the mean response) for houses with a living area of 2000 square feet falls within this range.
